### 2018-01-29 submission/manuscript.Rmd
add nick edits -- methods

***Human and Mock Samples:***
A single fecal sample was obtained from 4 individuals who were part of the Enterics Research Investigational Network (ERIN) and the processing and storage of these samples have been published previously [@erin_seekatz_2016]. Clinical data and other types of meta data were not utilized or accessed for this study. All samples were extracted using the MOBIO^TM^ PowerMag Microbiome RNA/DNA extraction kit (now Qiagen, MD, USA). The ZymoBIOMICS^TM^ Microbial Community DNA Standard (Zymo, CA, USA) was used in this study and is made up of *Pseudomonas aeruginosa*, *Escherichia coli*, *Salmonella enterica*, *Lactobacillus fermentum*, *Enterococcus faecalis*, *Staphylococcus aureus*, *Listeria monocytogenes*, and *Bacillus subtilis* at equal genomic DNA abundance  (http://www.zymoresearch.com/microbiomics/microbial-standards/zymobiomics-microbial-community-standards).

***PCR Protocol:***
The five different high fidelity (HiFi) Taq DNA polymerase that were tested were AccuPrime^TM^ (ThermoFisher, MA, USA), KAPA HIFI (Roche, IN, USA), Phusion (ThermoFisher, MA, USA), Platinum (ThermoFisher, MA, USA), and Q5 (New England Biolabs, MA, USA). The PCR cycle conditions for Platinum and Accuprime followed a previously published protocol [@protocol_Kozich_2013] (https://github.com/SchlossLab/MiSeq_WetLab_SOP/blob/master/MiSeq_WetLab_SOP_v4.md). If the HiFi Taq had a specific activation time that was different then 2 minutes, the specified activation was used instead. For Kappa and Q5 the protocol previously published by Gohl and colleagues was used [@taq_Gohl_2016]. For Phusion the company defined conditions were used but the same extension time as that used for Accuprime and Platinum was used.

The 30 cycle default was used but the cycle conditions started at 15 and increased by 5 up to 35 cycles and was used for both fecal and mock samples. The fecal PCR consisted of all 4 samples at 15, 20, 25, 30, and 35 cycles for each Taq (total samples = 100). Although, the mock communities also had 4 replicates used for 15, 20, 25, and 35 cycles, 10 replicates were used for 30 cycles for all Taq (total samples = 130). For all the mock community samples there was not enough PCR product at 15 cycles for adequate sequencing.

***Sequence Processing:***
The mothur software program was utilized for all sequence processing steps [@mothur_schloss_2009]. The protocol followed was similar to what has been previously publsihed [@protocol_Kozich_2013] (https://www.mothur.org/wiki/MiSeq_SOP). Two major differences from the stated protocol were the use of VSEARCH instead of UCHIME for chimera detection and the use of the OptiClust algorithmn instead of average neighbor for Operational Taxonomic Unit (OTU) generation at 97% similarity [@vsearch_Rognes_2016; @uchime_Edgar_2011; @opticlust_Westcott_2017]. Sequence error was determined using the seq.error command on mock samples after chimera removal and classification to the RDP to remove non-bacterial sequences [@mothur_schloss_2009; @vsearch_Rognes_2016; @rdp_Cole_2013].

***Statistical Analysis:***
All analysis was done with the R (v `r paste(version$major, ".", version$minor, sep = "")`) software package [@r_citation_2017]. Data tranformation and graphing was completed using the tidyverse package (v `r paste(packageVersion("tidyverse")[1])`) and colors selected using the viridis package (v `r paste(packageVersion("viridis")[1])`) [@tidyverse_2017; @viridis_citation_2017]. Differences in the total number of OTUs were analyzed using an ANOVA with a tukey post-hoc test. For the fecal samples the data was normalized to each individual by cycle number, to account for the biological variation between different people. Bray-Curtis matrices were generated using mothur after 100 sub-samplings at 1000, 5000, 10000, and 15000. The distance matrix data was analyzed using PERMANOVA with the vegan package (v `r paste(packageVersion("vegan")[1])`) [@vegan_citation] and kruskal-wallis tests within R. For both error and chimera analysis, samples were tested using Kruskal-Wallis with a Dunns post-hoc test. Where applicable correction for multiple comparison utilized the Benjamini-Hochberg method [@benjamini_controlling_1995].

***Analysis Workflow:***
The total number of OTUs after sub-sampling was analyzed for both the fecal and mock community samples. Cycle dependent affects on Bray-Curits indicies were next assessed for the fecal samples looking at both overall cycle differences and within individual differences for the previous cycle (e.g. 20 versus 25, 25 versus 30, etc.). Based on these observations we wanted to next analyze potential reasons as to why some of these differences may have occured. First, analysis of general sequence error rate, number of sequences with an error, and base subsitution were assessed in the mock community for each DNA polymerase. After assessing these errors, the total number of chimeras was determined after sequence processing. For the community based measures, the fecal samples were analyzed at 4 different sub-sampling levels, 1000, 5000, 10000, and 15000 while the mock community samples were analysed at 3 levels, 1000, 5000, 10000.

>>>>

***Human and Mock Samples:***
A single fecal sample was obtained from 4 individuals who were part of the Enterics Research Investigational Network (ERIN). The processing and storage of these samples have been published previously [@erin_seekatz_2016]. Clinical data and other types of meta data were not utilized or accessed for this study. All samples were extracted using the MOBIO^TM^ PowerMag Microbiome RNA/DNA extraction kit (now Qiagen, MD, USA). The ZymoBIOMICS^TM^ Microbial Community DNA Standard (Zymo, CA, USA) was used in this study and is made up of *Pseudomonas aeruginosa*, *Escherichia coli*, *Salmonella enterica*, *Lactobacillus fermentum*, *Enterococcus faecalis*, *Staphylococcus aureus*, *Listeria monocytogenes*, and *Bacillus subtilis* at equal genomic DNA abundance  (http://www.zymoresearch.com/microbiomics/microbial-standards/zymobiomics-microbial-community-standards).

***PCR Protocol:***
The five different high fidelity (HiFi) Taq DNA polymerase that were tested were AccuPrime^TM^ (ThermoFisher, MA, USA), KAPA HIFI (Roche, IN, USA), Phusion (ThermoFisher, MA, USA), Platinum (ThermoFisher, MA, USA), and Q5 (New England Biolabs, MA, USA). The PCR cycle conditions for Platinum and Accuprime followed a previously published protocol [@protocol_Kozich_2013] (https://github.com/SchlossLab/MiSeq_WetLab_SOP/blob/master/MiSeq_WetLab_SOP_v4.md). The HiFi DNA polymerase activation time was 2 minutes, unless a different activation was specified. For Kappa and Q5, the protocol previously published by Gohl and colleagues was used [@taq_Gohl_2016]. For Phusion, the company defined conditions were used except for extension time, where the Accuprime and Platinum settings were used.

Both fecal and mock samples cycle conditions started at 15 and increased by 5 up to 35 cycles with amplicons used at each 5-step increase for sequencing. The fecal PCR consisted of all 4 samples at 15, 20, 25, 30, and 35 cycles for each Taq (total samples = 100). Although, the mock communities also had 4 replicates for 15, 20, 25, and 35 cycles and 10 replicates for 30 cycles for all HiFi DNA polymerase (total samples = 130). No mock community sample had enough PCR product at 15 cycles for adequate 16S rRNA gene sequencing.

***Sequence Processing:***
The mothur software program was utilized for all sequence processing steps [@mothur_schloss_2009]. Generally, the protocol followed what has been previously publsihed [@protocol_Kozich_2013] (https://www.mothur.org/wiki/MiSeq_SOP). Two major differences from the stated protocol were the use of VSEARCH instead of UCHIME for chimera detection and the use of the OptiClust algorithmn instead of average neighbor for Operational Taxonomic Unit (OTU) generation at 97% similarity [@vsearch_Rognes_2016; @uchime_Edgar_2011; @opticlust_Westcott_2017]. Sequence error was determined using the seq.error command on mock samples after chimera removal and classification to the RDP to remove non-bacterial sequences [@mothur_schloss_2009; @vsearch_Rognes_2016; @rdp_Cole_2013].

***Statistical Analysis:***
All analysis was done with the R (v `r paste(version$major, ".", version$minor, sep = "")`) software package [@r_citation_2017]. Data tranformation and graphing was completed using the tidyverse package (v `r paste(packageVersion("tidyverse")[1])`) and colors selected using the viridis package (v `r paste(packageVersion("viridis")[1])`) [@tidyverse_2017; @viridis_citation_2017]. Differences in the total number of OTUs were analyzed using an ANOVA with a tukey post-hoc test. For the fecal samples the data was normalized to each individual by cycle number to account for the biological variation between people. Bray-Curtis matrices were generated using mothur after 100 sub-samplings at 1000, 5000, 10000, and 15000. The distance matrix data was analyzed using PERMANOVA with the vegan package (v `r paste(packageVersion("vegan")[1])`) [@vegan_citation] and kruskal-wallis tests within R. For both error and chimera analysis, samples were tested using Kruskal-Wallis with a Dunns post-hoc test. Where applicable correction for multiple comparison utilized the Benjamini-Hochberg method [@benjamini_controlling_1995].

***Analysis Workflow:***
The total number of OTUs after sub-sampling was analyzed for both the fecal and mock community samples. Cycle dependent affects on Bray-Curits indicies were next assessed for the fecal samples looking at both overall cycle differences and within individual differences for the previous cycle (e.g. 20 versus 25, 25 versus 30, etc.). Based on these observations we wanted to next analyze potential reasons for these differences. First, analysis of general sequence error rate, number of sequences with an error, and base subsitution were assessed in the mock community for each DNA polymerase. After assessing these errors, the total number of chimeras was determined after sequence processing. For the community based measures, the fecal samples were analyzed at 4 different sub-sampling levels (1000, 5000, 10000, and 15000) while the mock community samples were analysed at 3 levels (1000, 5000, 10000).

### 2018-01-29 submission/manuscript.tex
add nick edits -- methods

\textbf{\emph{Human and Mock Samples:}} A single fecal sample was
obtained from 4 individuals who were part of the Enterics Research
Investigational Network (ERIN) and the processing and storage of these
samples have been published previously (Seekatz et al., 2016). Clinical
data and other types of meta data were not utilized or accessed for this
study. All samples were extracted using the MOBIO\textsuperscript{TM}
PowerMag Microbiome RNA/DNA extraction kit (now Qiagen, MD, USA). The
ZymoBIOMICS\textsuperscript{TM} Microbial Community DNA Standard (Zymo,
CA, USA) was used in this study and is made up of \emph{Pseudomonas
aeruginosa}, \emph{Escherichia coli}, \emph{Salmonella enterica},
\emph{Lactobacillus fermentum}, \emph{Enterococcus faecalis},
\emph{Staphylococcus aureus}, \emph{Listeria monocytogenes}, and
\emph{Bacillus subtilis} at equal genomic DNA abundance
(\url{http://www.zymoresearch.com/microbiomics/microbial-standards/zymobiomics-microbial-community-standards}).

\textbf{\emph{PCR Protocol:}} The five different high fidelity (HiFi)
Taq DNA polymerase that were tested were AccuPrime\textsuperscript{TM}
(ThermoFisher, MA, USA), KAPA HIFI (Roche, IN, USA), Phusion
(ThermoFisher, MA, USA), Platinum (ThermoFisher, MA, USA), and Q5 (New
England Biolabs, MA, USA). The PCR cycle conditions for Platinum and
Accuprime followed a previously published protocol (Kozich et al., 2013)
(\url{https://github.com/SchlossLab/MiSeq_WetLab_SOP/blob/master/MiSeq_WetLab_SOP_v4.md}).
If the HiFi Taq had a specific activation time that was different then 2
minutes, the specified activation was used instead. For Kappa and Q5 the
protocol previously published by Gohl and colleagues was used (Gohl et
al., 2016). For Phusion the company defined conditions were used but the
same extension time as that used for Accuprime and Platinum was used.

The 30 cycle default was used but the cycle conditions started at 15 and
increased by 5 up to 35 cycles and was used for both fecal and mock
samples. The fecal PCR consisted of all 4 samples at 15, 20, 25, 30, and
35 cycles for each Taq (total samples = 100). Although, the mock
communities also had 4 replicates used for 15, 20, 25, and 35 cycles, 10
replicates were used for 30 cycles for all Taq (total samples = 130).
For all the mock community samples there was not enough PCR product at
15 cycles for adequate sequencing.

\textbf{\emph{Sequence Processing:}} The mothur software program was
utilized for all sequence processing steps (Schloss et al., 2009). The
protocol followed was similar to what has been previously publsihed
(Kozich et al., 2013) (\url{https://www.mothur.org/wiki/MiSeq_SOP}). Two
major differences from the stated protocol were the use of VSEARCH
instead of UCHIME for chimera detection and the use of the OptiClust
algorithmn instead of average neighbor for Operational Taxonomic Unit
(OTU) generation at 97\% similarity (Edgar et al., 2011; Rognes et al.,
2016; Westcott \& Schloss, 2017). Sequence error was determined using
the seq.error command on mock samples after chimera removal and
classification to the RDP to remove non-bacterial sequences (Schloss et
al., 2009; Cole et al., 2013; Rognes et al., 2016).

\textbf{\emph{Statistical Analysis:}} All analysis was done with the R
(v 3.4.3) software package (R Core Team, 2017). Data tranformation and
graphing was completed using the tidyverse package (v 1.2.1) and colors
selected using the viridis package (v 0.4.1) (Garnier, 2017; Wickham,
2017). Differences in the total number of OTUs were analyzed using an
ANOVA with a tukey post-hoc test. For the fecal samples the data was
normalized to each individual by cycle number, to account for the
biological variation between different people. Bray-Curtis matrices were
generated using mothur after 100 sub-samplings at 1000, 5000, 10000, and
15000. The distance matrix data was analyzed using PERMANOVA with the
vegan package (v 2.4.5) (Oksanen et al., 2017) and kruskal-wallis tests
within R. For both error and chimera analysis, samples were tested using
Kruskal-Wallis with a Dunns post-hoc test. Where applicable correction
for multiple comparison utilized the Benjamini-Hochberg method
(Benjamini \& Hochberg, 1995).

\textbf{\emph{Analysis Workflow:}} The total number of OTUs after
sub-sampling was analyzed for both the fecal and mock community samples.
Cycle dependent affects on Bray-Curits indicies were next assessed for
the fecal samples looking at both overall cycle differences and within
individual differences for the previous cycle (e.g.~20 versus 25, 25
versus 30, etc.). Based on these observations we wanted to next analyze
potential reasons as to why some of these differences may have occured.
First, analysis of general sequence error rate, number of sequences with
an error, and base subsitution were assessed in the mock community for
each DNA polymerase. After assessing these errors, the total number of
chimeras was determined after sequence processing. For the community
based measures, the fecal samples were analyzed at 4 different
sub-sampling levels, 1000, 5000, 10000, and 15000 while the mock
community samples were analysed at 3 levels, 1000, 5000, 10000.

>>>>

\textbf{\emph{Human and Mock Samples:}} A single fecal sample was
obtained from 4 individuals who were part of the Enterics Research
Investigational Network (ERIN). The processing and storage of these
samples have been published previously (Seekatz et al., 2016). Clinical
data and other types of meta data were not utilized or accessed for this
study. All samples were extracted using the MOBIO\textsuperscript{TM}
PowerMag Microbiome RNA/DNA extraction kit (now Qiagen, MD, USA). The
ZymoBIOMICS\textsuperscript{TM} Microbial Community DNA Standard (Zymo,
CA, USA) was used in this study and is made up of \emph{Pseudomonas
aeruginosa}, \emph{Escherichia coli}, \emph{Salmonella enterica},
\emph{Lactobacillus fermentum}, \emph{Enterococcus faecalis},
\emph{Staphylococcus aureus}, \emph{Listeria monocytogenes}, and
\emph{Bacillus subtilis} at equal genomic DNA abundance
(\url{http://www.zymoresearch.com/microbiomics/microbial-standards/zymobiomics-microbial-community-standards}).

\textbf{\emph{PCR Protocol:}} The five different high fidelity (HiFi)
Taq DNA polymerase that were tested were AccuPrime\textsuperscript{TM}
(ThermoFisher, MA, USA), KAPA HIFI (Roche, IN, USA), Phusion
(ThermoFisher, MA, USA), Platinum (ThermoFisher, MA, USA), and Q5 (New
England Biolabs, MA, USA). The PCR cycle conditions for Platinum and
Accuprime followed a previously published protocol (Kozich et al., 2013)
(\url{https://github.com/SchlossLab/MiSeq_WetLab_SOP/blob/master/MiSeq_WetLab_SOP_v4.md}).
The HiFi DNA polymerase activation time was 2 minutes, unless a
different activation was specified. For Kappa and Q5, the protocol
previously published by Gohl and colleagues was used (Gohl et al.,
2016). For Phusion, the company defined conditions were used except for
extension time, where the Accuprime and Platinum settings were used.

Both fecal and mock samples cycle conditions started at 15 and increased
by 5 up to 35 cycles with amplicons used at each 5-step increase for
sequencing. The fecal PCR consisted of all 4 samples at 15, 20, 25, 30,
and 35 cycles for each Taq (total samples = 100). Although, the mock
communities also had 4 replicates for 15, 20, 25, and 35 cycles and 10
replicates for 30 cycles for all HiFi DNA polymerase (total samples =
130). No mock community sample had enough PCR product at 15 cycles for
adequate 16S rRNA gene sequencing.

\textbf{\emph{Sequence Processing:}} The mothur software program was
utilized for all sequence processing steps (Schloss et al., 2009).
Generally, the protocol followed what has been previously publsihed
(Kozich et al., 2013) (\url{https://www.mothur.org/wiki/MiSeq_SOP}). Two
major differences from the stated protocol were the use of VSEARCH
instead of UCHIME for chimera detection and the use of the OptiClust
algorithmn instead of average neighbor for Operational Taxonomic Unit
(OTU) generation at 97\% similarity (Edgar et al., 2011; Rognes et al.,
2016; Westcott \& Schloss, 2017). Sequence error was determined using
the seq.error command on mock samples after chimera removal and
classification to the RDP to remove non-bacterial sequences (Schloss et
al., 2009; Cole et al., 2013; Rognes et al., 2016).

\textbf{\emph{Statistical Analysis:}} All analysis was done with the R
(v 3.4.3) software package (R Core Team, 2017). Data tranformation and
graphing was completed using the tidyverse package (v 1.2.1) and colors
selected using the viridis package (v 0.4.1) (Garnier, 2017; Wickham,
2017). Differences in the total number of OTUs were analyzed using an
ANOVA with a tukey post-hoc test. For the fecal samples the data was
normalized to each individual by cycle number to account for the
biological variation between people. Bray-Curtis matrices were generated
using mothur after 100 sub-samplings at 1000, 5000, 10000, and 15000.
The distance matrix data was analyzed using PERMANOVA with the vegan
package (v 2.4.5) (Oksanen et al., 2017) and kruskal-wallis tests within
R. For both error and chimera analysis, samples were tested using
Kruskal-Wallis with a Dunns post-hoc test. Where applicable correction
for multiple comparison utilized the Benjamini-Hochberg method
(Benjamini \& Hochberg, 1995).

\textbf{\emph{Analysis Workflow:}} The total number of OTUs after
sub-sampling was analyzed for both the fecal and mock community samples.
Cycle dependent affects on Bray-Curits indicies were next assessed for
the fecal samples looking at both overall cycle differences and within
individual differences for the previous cycle (e.g.~20 versus 25, 25
versus 30, etc.). Based on these observations we wanted to next analyze
potential reasons for these differences. First, analysis of general
sequence error rate, number of sequences with an error, and base
subsitution were assessed in the mock community for each DNA polymerase.
After assessing these errors, the total number of chimeras was
determined after sequence processing. For the community based measures,
the fecal samples were analyzed at 4 different sub-sampling levels
(1000, 5000, 10000, and 15000) while the mock community samples were
analysed at 3 levels (1000, 5000, 10000).
